Bayesian Hyper-Heuristics

### 2023-01-22 article/sections/01-introduction.tex
🚧 (article): Section 2 done, WIP

Over the past few years, modern hardware capabilities have improved to the point where workloads in the field of \acs{ML} that were previously computationally infeasible, are now possible. One such sub-field of \acs{ML} is \acp{ANN}.

The performance and capabilities of \acp{ANN} is largely influenced by the learning process used. As such, a popular field of focus for studying \acp{FFNN} is the process by which these models are trained. \acp{ANN} are trained by optimisation algorithms known as heuristics.

Many different heuristics have been developed and used to train \acp{FFNN}~\cite{ref:gudise:2003, ref:rakitianskaia:2012, ref:montana:1989}. Each of these heuristics have different search behaviours, characteristics, strengths and weaknesses. Finding the best heuristic to train the \acs{FFNN} is required in order to yield optimal results. This process is often non-trivial and could be a time-consuming exercise.  Consider that selection of the best heuristic as applied to optimisation problems, such as training \acp{FFNN}, is problem specific~\cite{ref:allen:1996, ref:drake:2020, ref:pillay:2018}.

A recent suggestion related to the field of \index{meta-learning}meta-learning is to dynamically select and/or adjust the heuristic used throughout the training process. This approach focuses on the hybridisation of learning paradigms.

One such form of hybridisation of learning paradigms is that of hybridisation of different \textit{heuristics} as they are applied to some optimisation problem~\cite{ref:burke:2013}. These methods are referred to as \acfp{HH} and focus on finding the best heuristic in \textit{heuristic space} to solve a specific problem.

In the general context of optimisation, many different types of \acp{HH} have been implemented and applied to many different problems. Some notable examples include the \index{simulated annealing}simulated annealing-based \acs{HH} by~\citeauthor{ref:dowsland:2007}~\cite{ref:dowsland:2007}, the \index{tabu-search}tabu-search \acs{HH} of~\citeauthor{ref:burke:2010}~\cite{ref:burke:2010}, the \index{heterogeneous meta-hyper-heuristic}heterogeneous meta-hyper-heuristic by~\citeauthor{ref:grobler:2012}~\cite{ref:grobler:2012} and work done by~\citeauthor{ref:vanderstockt:2018}~\cite{ref:vanderstockt:2018} on the analysis of selection hyper-heuristics for population-based \acp{MH} in real-valued dynamic optimization. However, research on the application of \acp{HH} in the context of \acs{FFNN} training is scarce.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} provides the first research in this field, applying a \acs{HH} to \acs{FFNN} training.

@@

The remainder of this article is structured as follows: Section~\ref{sec:anns} provides background information on \acp{ANN}. The focus is on training of \acp{FFNN}. Section~\ref{sec:heuristics} provides details on various types of heuristics and \acp{MH} that have been used to train \acp{FFNN}. Section~\ref{sec:hhs} presents background information on the details of \acp{HH} and \index{meta-learning}meta-learning in general. Section~\ref{sec:probability} presents background information on probability theory. Section~\ref{sec:bhh} presents the developed \Acs{BHH}. Section~\ref{sec:methodology} presents a detailed description of the empirical process and the setup of each experiment. Section~\ref{sec:results} provides and discusses the results of the empirical study in detail. Section~\ref{sec:conclusion} summarises the research done in this dissertation along with a brief overview of the findings made throughout the research process.

>>>>

Over the past few years, modern hardware capabilities have improved to the point where workloads in the field of \acs{ML} that were previously computationally infeasible, are now possible. One such sub-field of \acs{ML} is \acp{ANN}. The performance and capabilities of \acp{ANN} is largely influenced by the learning process used. As such, a popular field of focus for studying \acp{ANN} is the process by which these models are trained. \acp{ANN} are trained by optimisation algorithms known as heuristics.

Many different heuristics have been developed and used to train \acp{ANN}~\cite{ref:gudise:2003, ref:rakitianskaia:2012, ref:montana:1989}. Each of these heuristics have different search behaviours, characteristics, strengths and weaknesses. Finding the best heuristic to train the \acs{ANN} is required in order to yield optimal results. This process is often non-trivial and could be a time-consuming exercise.  Selection of the best heuristic to train \acp{ANN}, is often problem specific~\cite{ref:allen:1996, ref:drake:2020, ref:pillay:2018}.

A recent suggestion related to the field of \index{meta-learning}meta-learning is to dynamically select and/or adjust the heuristic used throughout the training process. This approach focuses on the hybridisation of learning paradigms. One such form of hybridisation of learning paradigms is that of hybridisation of different \textit{heuristics} as they are applied to some optimisation problem~\cite{ref:burke:2013}. These methods are referred to as \acfp{HH} and focus on finding the best heuristic in \textit{heuristic space} to solve a specific problem.

In the general context of optimisation, many different types of \acp{HH} have been implemented and applied to many different problems. However, research on the application of \acp{HH} in the context of \acs{ANN} training is scarce. \citeauthor{ref:nel:2021}~\cite{ref:nel:2021} provides the first research in this field, applying a \acs{HH} to \acs{FFNN} training.

@@

The remainder of this article is structured as follows: Section~\ref{sec:anns} provides background information on \acp{ANN}. Section~\ref{sec:heuristics} provides details on various types of heuristics that have been used to train \acp{FFNN}. Section~\ref{sec:hhs} presents background information on \acp{HH} and \index{meta-learning}meta-learning. Section~\ref{sec:probability} presents background information on probability theory. Section~\ref{sec:bhh} presents the developed \Acs{BHH}. Section~\ref{sec:methodology} presents a detailed description of the empirical process and the setup of each experiment. Section~\ref{sec:results} provides and discusses the results of the empirical study. Section~\ref{sec:conclusion} summarises the research that is done along with a brief overview of the findings made throughout the research process.

### 2023-01-22 article/sections/02-artificial-neural-networks.tex
🚧 (article): Section 2 done, WIP

>>>>

\acp{ANN} can generally be described as well-organised structures of mathematical computation and are inspired by the biological brain~\cite{ref:engelbrecht:2007}. \acp{ANN} have been successfully applied to a range of problem classes.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} summarises some common problems that are solved using \acp{ANN}. These include classification~\cite{ref:khan:2001}, pattern matching~\cite{ref:cannady:1998, ref:kumar:1994}, pattern completion~\cite{ref:dayhoff:2001}, optimisation~\cite{ref:specht:1991} and data mining~\cite{ref:singh:2009}.

\subsection{Artificial Neuron}
\label{sec:anns:an}

\Acp{AN} provide the fundamental building blocks for \acp{ANN}. The \acs{AN} implements a non-linear mapping from $\mathbb{R}^{I}$ to $\mathbb{R}^{T}$, usually in the ranges $[0,1]$ or $[-1,1]$, depending on the problem being solved ~\cite{ref:engelbrecht:2007}. $I$ is the total number of dimensions of the input in real-number space ($\mathbb{R}$), and $T$ is the total number of dimensions of the target (desired output) in real-number space. The \acs{AN} implements various components and is illustrated in Figure~\ref{fig:artificial_neuron}.

\begin{figure}[htb]
	\centering
	\includegraphics[width=\textwidth]{artificial_neuron.pdf}
	\caption[The \index{artificial neuron}artificial neuron]{An illustration of the \index{artificial neuron}artificial neuron.}
	\label{fig:artificial_neuron}
\end{figure}

Each of the components of the \acs{AN}, presented in Figure~\ref{fig:artificial_neuron}, is inspired by some element of the \acs{BN}. The input ($\boldsymbol{x}$) models the activations received from the pre-synaptic neuron. The weights ($\boldsymbol{v}$), model the \index{synapse}synapses and connection strengths. The \index{net input signal}net input signal ($net$) models the net resulting activations from all connected pre-synaptic neurons. The biases ($\theta$), model a mechanism introduced to influence the strength of the activation (output signal) of the \acs{AN}. The activation function ($f$) models the action potential of the \acs{BN} and is based on the net input signal. Finally, the output ($\boldsymbol{y}$) models the activation by the post-synaptic neuron.

\subsection{Feedforward Neural Networks}
\label{sec:anns:ffnn}

Multiple \acp{AN} can be organised and used together forming a ``network'' of \acp{AN} referred to as an \acf{ANN}. The architecture of the \acs{ANN} refers to the way in which \acp{AN} are organised. \acp{ANN} can be organised in layers where a single layer can contain multiple \acp{AN}. Generally, each layer makes use of the same \index{activation function}activation function. Output from one layer is propagated as input to the next layer. The topology of the \acs{ANN} refers to the way that layers of \acp{AN} are connected to each other. There are many different topologies~\cite{ref:miikkulainen:2010}.

This research focuses on a particular type of \acs{ANN}, referred to as \acp{FFNN}. \acp{FFNN} were the first and simplest type of \acp{ANN} developed~\cite{ref:schmidhuber:2015} and implement an architecture consisting of input, hidden and output layers by arranging them in sequential order. Furthermore, \acp{FFNN} implement \index{fully connected topology}fully connected topologies, where each \acs{AN} in one layer is connected to all \acp{AN} in the next, without any cycles~\cite{ref:zell:1994}.

In \acp{FFNN}, information moves forward, in one direction, from the input nodes, through the hidden nodes and finally to the output nodes. An illustration of a \acs{FFNN} is given in Figure~\ref{fig:ffnn}.

\begin{figure}[htb]
	\centering
	\includegraphics[width=0.98\textwidth]{feedforward_neural_network.pdf}
	\caption[A \index{feedforward neural network}feedforward neural network]{An illustration of a \index{feedforward neural network}feedforward neural network implementing input, hidden and output layers using a fully connected topology.}
	\label{fig:ffnn}
\end{figure}

In Figure~\ref{fig:ffnn}, $x_i$ refers to the $i$-th dimension in the input vector $\boldsymbol{x}$, $h_j$ refers to the $j$-th dimension in the hidden layer, $y_k$ refers to the $k$-th dimension in the output vector $\boldsymbol{y}$, $v_{i,j}$ refers to the weight associated with input node $x_i$ and the hidden node $h_j$, and $w_{j,k}$ refers to the weight associated with hidden node $h_j$ and the output node $y_k$.

\subsection{Training}
\label{sec:anns:training}

\textit{Training} is the process whereby the weights of the \acs{FFNN} are systematically changed with the aim of improving the \textit{performance} of the \acs{FFNN}. Finding the optimal weights that produce the best performance on a given task is an optimisation problem. The optimisation algorithm used to find the optimal weights is referred to as a \index{heuristic}\textit{heuristic}. \index{heuristic}Heuristics search for possible solutions in the solution-space and make use of information from the search space to guide to process.

During the training process, the \acs{FFNN} is exposed to data while trying to produce some target outcome. The degree to which the produced outcome differs from the target outcome is referred to as \textit{loss}. Since training of \acp{FFNN} is an optimisation problem, the goal of the training process is to minimise the loss. The loss is calculated using an error function. This research focuses on a number of error functions, including \acf{SSE}, \acf{MSE}, \acf{RMSE}, \acf{MAE}, \acf{BinXE}, \acf{CatXE} and \acf{SparseCatXE}.

\index{supervised learning}Supervised learning is the process of training where the data that is presented to the \acs{FFNN} during training, includes the desired solution~\cite{ref:geron:2017}. The \acs{FFNN} learns the mapping function from the input to the target output~\cite{ref:brownlee:2016}.

### 2023-01-22 article/sections/03-heuristics.tex
🚧 (article): Section 2 done, WIP

\chapter{Heuristics}\label{chap:heuristics}

\begin{quotation}
	``It is not the strongest of the species that survives. It is also not the most intelligent that survives. It is the one that is the most adaptable to change.''
\end{quotation}
\begin{flushright}
	- Charles Darwin
\end{flushright}

Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. Finding the best technique to use to train a \acs{FFNN} has been shown to be problem dependent in many cases~\cite{ref:kheiri:2017}. Every technique has its characteristics, constraints, advantages and disadvantages. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. Gradient-based techniques are not without flaws and can, for example, yield slow convergence or get trapped in local optima~\cite{ref:mingguang:2009}. Other techniques have also been used to successfully train \acp{FFNN}, including \acp{MH} such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}.

Chapter~\ref{chap:anns} briefly introduced the reader to the concept of \index{heuristic}\index{heuristic}heuristics and \acfp{MH}s. This chapter presents more detailed background information on various different \index{heuristic}heuristics that have been used to train \acp{FFNN}. Broadly speaking, this chapter focuses on two different groups of \index{heuristic}heuristics, including classical gradient-based approaches and population-based \acp{MH}. Each technique is presented and discussed in detail. Pseudo-code algorithms are provided for each technique and discussions follow on advantages, disadvantages, capabilities and limitations. The remainder of this chapter is structured as follows:

\begin{itemize}
	\item \textbf{Section~\ref{sec:heuristics:optimisation}} provides a brief review of optimisation. It is shown that training of \acp{FFNN} is an optimisation problem.

	\item \textbf{Section~\ref{sec:heuristics:what_is_a_heuristic}} provides background information on the origins and definition of the term \index{heuristic}\textit{heuristic}. It is shown that \index{heuristic}heuristics are a class of algorithms that are used to solve optimisation problems.

	\item \textbf{Section~\ref{sec:heuristics:gd}} presents seven low-level, gradient-based \index{heuristic}heuristics, including \acf{SGD}, \acf{Momentum}, \acf{NAG}, \acf{Adagrad}, \acf{RMSProp}, \acf{Adadelta} and \acf{Adam}.

	\item \textbf{Section~\ref{sec:heuristics:mh}} presents three different population-based \acp{MH}, including \acf{PSO}, \acf{DE} and \acfp{GA}.

	\item \textbf{Section~\ref{sec:heuristics:summary}} provides a brief summary of the chapter.
\end{itemize}

\section{Optimisation}\label{sec:heuristics:optimisation}

Optimisation is the task of finding a solution to a given problem that is better than alternative solutions. Better stated by \citeauthor{ref:oldewage:2017}~\cite{ref:oldewage:2017}, optimisation is the task of finding values for a set of variables such that some measure of optimality is satisfied given a set of constraints. \citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} breaks optimisations problems down into three components:

\begin{itemize}
	\item An \textbf{objective function}: Represents the quantity to be optimised and is used as the ``measure of optimality''. Optimisation can be defined in terms of the minimisation or maximisation of the objective function $f$.

	\item A \textbf{set of unknowns or independent variables}: Affects the outcome of the objective function $f$ and is denoted as $x$. $f(x)$ is thus the quantification of the objective function over the unknowns, represented by $x$. Note that $x$ could be a scalar value, a vector or a matrix and notation is left out for simplicity.

	\item A \textbf{set of constraints}: Restrict and limit the values that can be assigned to the unknowns, represented by $x$. Optimisation problems that must adhere to a set of constraints are referred to as \acfp{CSP}.
\end{itemize}

Optimisation problems come in a wide variety, and can be defined in terms of the number of variables used (uni- vs. multivariate), the number of objective functions used (single- vs. multi-objective), the degree of linearity (linear vs. quadratic/polynomial), the number of optima (uni- vs. multi-modal), the nature of the environment (static vs. dynamic), the types of variables used (separable vs. inseparable, discrete vs. continuous) and the set of constraints that the solution must adhere to (constrained vs unconstrained).

Optima can be defined as \textit{local} or \textit{global} optima. Local optima is the best optimisation of $f(x)$ in a neighbourhood of solutions, while the global optima is the best optimisation of $f(x)$ over all solutions in the solution space.

As stated in Chapter~\ref{chap:anns}, the training of an \acp{FFNN} is a particular type of optimisation problem, where the goal is to find the configuration of \textit{weights}, such that the \acs{FFNN} yields output that minimises some loss function. The mechanism by which the optimal weights for a \acs{FFNN} is sought out, is executed by an \index{optimisation algorithm}optimisation algorithm known as a \index{heuristic}heuristic.

\section{What is a heuristic?}\label{sec:heuristics:what_is_a_heuristic}

The term \textit{heuristic} comes from the Latin word \textit{heuristicus} which means ``to find out or discover''.~\citeauthor{ref:romanycia:1985}~\cite{ref:romanycia:1985} provide a complete study on the history and origins of the term \index{heuristic}\textit{heuristic}. From their research, a proposal is made to define \index{heuristic}heuristics in the context of \acf{AI}, as any device, be it a program, rule, piece of knowledge, which is added to a problem-solving system, in expectation that, on average, the performance will improve.

In the context of this dissertation, a \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Different \index{heuristic}heuristics make use of different information during the search process~\cite{ref:kheiri:2017}. During training of \acp{FFNN}, \index{heuristic}heuristics such as gradient-based \index{heuristic}heuristics make use of the derivatives obtained by evaluating the \acs{FFNN}. It can thus be said that gradient-based \index{heuristic}heuristics make use of information directly from the \textit{search space}. On the contrary, \index{heuristic}heuristics such as \acp{MH} make use of meta-information obtained as a result of evaluating the \acs{FFNN}~\cite{ref:blum:2003}. The meta-information that \acp{MH} make use of could include ranked-performance of a population of candidate solutions, referred to as \textit{entities}. \index{meta-heuristic}Meta-heuristics are useful when there is imperfect information about the search space~\cite{ref:bianchi:2009}, and are generally less problem-specific than other classes of \index{heuristic}heuristics~\cite{ref:blum:2003}. This dissertation takes a particular interest in gradient-based \index{heuristic}heuristics and \acp{MH}.

\section{Gradient-Based Heuristics}\label{sec:heuristics:gd}

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the model being optimised. Specifically, in the context of a minimisation problem, these techniques are called \acf{GD} heuristics as they \textit{minimise} some loss function. \Acs{GD} is generally attributed to Cauchy~\cite{ref:lemarechal:2012}, who first suggested it in 1847. In 1907 Hadamard~\cite{ref:hadamard:1908} independently proposed a similar method.

Although gradient-based \index{heuristic}heuristics where not the first \index{heuristic}heuristics used to train \acp{FFNN}
\cite{ref:engelbrecht:2007}, they are certainly the most widely used. Gradient-based heuristics have become increasingly popular partly due to their simplicity and low computational overhead compared to other \index{heuristic}heuristics such as \acp{MH} and other second-order derivative methods such as Newtons' method. There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acf{GD} framework called \acf{BP}.

\subsection{Backpropagation}\label{sec:heuristics:gd:backpropagation}

Chapter~\ref{chap:anns} introduced \index {supervised learning}supervised learning and presented a number of loss functions. In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \acs{FFNN} in order to minimise the error~\cite{ref:engelbrecht:2007}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient. An illustration of \acs{GD} is given in Figure~\ref{fig:heuristics:gd:gd_illustration}.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.75\textwidth]{images/gradient_descent.pdf}
	\caption{An illustration of \acf{GD} over various time steps showing the minimisation of the error with regards to weight value.}
	\label{fig:heuristics:gd:gd_illustration}
\end{figure}

In the context of training \textit{shallow} \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer to the input layer, updating the weights at the output and hidden layers. The algorithm that propagates the error signal backwards is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\mathcal{E}$, with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps:

@@

The update step for \acs{GD} by means of \acs{BP} can be formulated as is shown in Equations~\eqref{eq:heuristics:gd:update_step_part_1} to \eqref{eq:heuristics:gd:update_step_part_4}. The general weight update step is given as

\begin{equation}
	\label{eq:heuristics:gd:update_step_part_1}
	w_{i}(t) = w_{i}(t-1) + \Delta w_{i}(t)
\end{equation}

where $w_{i}$ is the weight value at index $i$, $t$ is the time step, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and time step $t$. The delta weight term $\Delta w_{i}(t)$, is given as

\begin{equation}
	\label{eq:heuristics:gd:update_step_part_2}
	\Delta w_{i}(t) = -\eta\frac{\partial \epsilon}{\partial w_{i}}
\end{equation}

where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each time step. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight vector $\boldsymbol{w}$, at index $i$, is given as

\begin{equation}
	\label{eq:heuristics:gd:update_step_part_3}
	\frac{\partial \epsilon}{\partial w_{i}} = -2(t_{i,p} - o_{i,p})\frac{\partial f}{\partial net_{p}}z_{i,p}
\end{equation}

where $t_{i,p}$ refers to the $i$-th element of the target value for pattern $p$, $o_{i,p}$ refers to the $i$-th element of the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

Finally, assuming the use of the \index{sigmoid}sigmoid activation function, the partial derivative of the activation function, relative to the net input signal is given as

\begin{equation}
	\label{eq:heuristics:gd:update_step_part_4}
	\frac{\partial f}{\partial net_{p}} = o_p(1 - o_{p})
\end{equation}

Again, assuming the use of \acs{SSE} as the loss function, the pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

\begin{algorithm}[htb]
	\caption{The pseudo-code algorithm for the generic \acf{GD} heuristic.}
	\label{algo:heuristics:gd:gd}
	\begin{algorithmic}
		\State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the time step/epoch $T=0$;
		\While{stopping conditions are not met}
		\State Let $\epsilon_{T} = 0$;
		\For{each training pattern p}
		\State Calculate the output of the hidden and output layers (feedforward);
		\State Compute the error signals of the hidden and output layers;
		\State Adjust the weights of the output and hidden layers (backpropagate);
		\State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
		\EndFor
		\State $T += 1$;
		\EndWhile
	\end{algorithmic}
\end{algorithm}

@@

\subsection{Stochastic vs. Batch Training}\label{sec:heuristics:gd:sgd}

This section shines light on the algorithmic implementation of \acs{BP} with specific context to stochastic and batch training. The implementation of \acs{GD} using stochastic training is referred to as \acf{SGD}.

\acs{SGD} was one of the first widely used heuristics to train \acs{FFNN}, however, it is not without flaws. With stochastic training, only one training pattern is presented at each iteration/epoch. As such, weight updates are done with high variance and noise~\cite{ref:ruder:2016}. An illustration of the fluctuations caused by \acs{SGD} during training is given in Figure~\ref{fig:heuristics:gd:sgd}.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.8\textwidth]{images/sgd.pdf}
	\caption{An illustration of \acf{SGD} fluctuations during training as taken from~\cite{ref:sgd:2006}.}
	\label{fig:heuristics:gd:sgd}
\end{figure}

With batch training, all the training patterns are presented at once and the \acs{FFNN} converges to the global minimum of the loss function w.r.t $\boldsymbol{w}$ (assuming the loss function is convex). However, this is computationally expensive and impractical to implement in many situations. While \acs{SGD} is able to jump out of local minima into new, potentially better local minima~\cite{ref:ruder:2016}, \acs{SGD} complicates convergence to the minima as \acs{SGD} can potentially keep overshooting better minima. By slowly decreasing the learning rate $\eta$ during training, the same convergence behaviour is achieved as with batch training.

A compromise is to make use of mini-batches of training patterns, where a small number of training patterns are presented to the \acs{FFNN} at once. This is referred to as \textit{mini-batch} training. Input patterns from mini-batches are approximations of the total population of training patterns. According to~\citeauthor{ref:ruder:2016}~\cite{ref:ruder:2016} this has two main advantages:

\begin{itemize}
	\item Mini-batch training reduces the high variance of weight updates as observed for \acs{SGD}, which leads to better convergence.

	\item Mini-batch training allows for the implementation of \acs{GD} using highly optimised matrix operations, common to the state-of-the-art \acs{ML} libraries used today.
\end{itemize}

@@

\subsection{Momentum}\label{sec:heuristics:gd:momentum}

@@

\subsection{Nesterov Accelerated Gradients}\label{sec:heuristics:nag}

@@

\subsection{Adaptive Gradients}\label{sec:heuristics:adagrad}

@@

\subsection{Adaptive Learning Rate}\label{sec:heuristics:adadelta}

@@

\subsection{Root Mean Squared Error Propagation}\label{sec:heuristics:rmsprop}

@@

\subsection{Adaptive Moments Estimation}\label{sec:heuristics:adam}

@@

\section{Meta-Heuristics}\label{sec:heuristics:mh}

@@

\subsection{Particle Swarm Optimisation}\label{sec:heuristics:mh:pso}

@@

\subsection{Differential Evolution}\label{sec:heuristics:mh:de}

@@

\subsection{Genetic Algorithms}\label{sec:heuristics:mh:ga}

@@

\section{Summary}
\label{sec:heuristics:summary}

This chapter provided detailed background information on \index{heuristic}heuristics. Different heuristics were presented that have been shown to be able to train \acp{FFNN}. Two main groups of \index{heuristic}heuristics where presented and discussed in detail. These include gradient-based \index{heuristic}heuristics and \acfp{MH}. A number of different variants for each group have been presented and was discussed in detail.

>>>>

\section{Heuristics}
\label{sec:heuristics}

A \index{heuristic}heuristic refers to an algorithmic search technique that serves as a guide to a search process where good solutions to a optimisation problem is being sought out. Many different techniques have been used to train \acp{FFNN}~\cite{ref:kingma:2014}. At the time of writing, the majority of work that is published around the training of \acp{FFNN}, involves the use of gradient-based techniques~\cite{ref:nel:2021}. However, other techniques have also been used to successfully train \acp{FFNN}, including \acp{MH} such as \acf{PSO}~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}, \acf{DE}~\cite{ref:espinal:2011} and \acfp{GA}~\cite{ref:gupta:1999}. This research takes a particular interest in both gradient-based heuristics as well as \acp{MH}.

\subsection{Gradient-Based Heuristics}\label{sec:heuristics:gd}

Gradient-based \index{heuristic}heuristics are optimisation techniques that make use of derivates obtained from evaluating the model being optimised. There are many variants of gradient-based \index{heuristic}heuristics, however, they all fundamentally apply the same generic \acf{GD} framework called \acf{BP}.

In the context of supervised learning, loss functions produce a scalar value $\epsilon$, that represents the error between the output of the \acs{FFNN} and the desired output. When using \acs{GD} to train \acp{FFNN}, the \index{loss function}loss function is used to adjust the weights of the \acs{FFNN} in order to minimise the error~\cite{ref:engelbrecht:2007}.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} states that training of \acp{FFNN} using \acs{GD}, is done by calculating the gradient of $\epsilon$ in \textit{weight-space}, and then moving the weight vector along the negative gradient.

In the context of training \acp{FFNN} using \index{supervised learning}supervised learning, the error signal is propagated backwards in the network from the output layer, through the hidden layer to the input layer, updating the weights at the output and hidden layers. The algorithm that propagates the error signal backwards is known as \acf{BP}. \Acs{BP} was popularised by~\citeauthor{ref:werbos:1994}~\cite{ref:werbos:1994}.~\citeauthor{ref:nel:2021}~\cite{ref:nel:2021} states that \acs{BP} provides a procedure for updating the network weights, layer by layer, by evaluating the derivatives of the error function $\mathcal{E}$, with respect to the weights at each layer.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} describes the \acs{BP} process in two steps:

@@

% The update step for \acs{GD} by means of \acs{BP} can be formulated as is shown in Equations~\eqref{eq:heuristics:gd:update_step_part_1} to \eqref{eq:heuristics:gd:update_step_part_4}. The general weight update step is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_1}
% 	w_{i}(t) = w_{i}(t-1) + \Delta w_{i}(t)
% \end{equation}

% where $w_{i}$ is the weight value at index $i$, $t$ is the time step, $\Delta w_{i}(t)$ is the weight update vector at index $i$ and time step $t$. The delta weight term $\Delta w_{i}(t)$, is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_2}
% 	\Delta w_{i}(t) = -\eta\frac{\partial \epsilon}{\partial w_{i}}
% \end{equation}

% where $\eta$ is the learning rate and $\frac{\partial \epsilon}{\partial w_{i}}$ is the \textit{gradient} of the error, as a result of a loss function $\mathcal{E}$, relative to the weight at index $i$. The learning rate controls the step-size that is taken in the direction of the negative gradient at each time step. Assuming the use of \acs{SSE} as the loss function, the partial derivative of $\epsilon$, relative to the weight vector $\boldsymbol{w}$, at index $i$, is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_3}
% 	\frac{\partial \epsilon}{\partial w_{i}} = -2(t_{i,p} - o_{i,p})\frac{\partial f}{\partial net_{p}}z_{i,p}
% \end{equation}

% where $t_{i,p}$ refers to the $i$-th element of the target value for pattern $p$, $o_{i,p}$ refers to the $i$-th element of the output value for pattern $p$, $f$ refers to the \index{activation function}activation function, $net_{p}$ refers to the \index{net input signal}net input signal for pattern $p$, and $z_{i,p}$ refers to the input value at index $i$ for pattern $p$.

% Finally, assuming the use of the \index{sigmoid}sigmoid activation function, the partial derivative of the activation function, relative to the net input signal is given as

% \begin{equation}
% 	\label{eq:heuristics:gd:update_step_part_4}
% 	\frac{\partial f}{\partial net_{p}} = o_p(1 - o_{p})
% \end{equation}

% Again, assuming the use of \acs{SSE} as the loss function, the pseudo-code implementation for the generic \acs{GD} algorithm is taken from~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} and is presented in Algorithm~\ref{algo:heuristics:gd:gd} below.

% \begin{algorithm}[htb]
% 	\caption{The pseudo-code algorithm for the generic \acf{GD} heuristic.}
% 	\label{algo:heuristics:gd:gd}
% 	\begin{algorithmic}
% 		\State Initialise the \acs{FFNN} weights and biases, the learning rate $\eta$ and the time step/epoch $T=0$;
% 		\While{stopping conditions are not met}
% 		\State Let $\epsilon_{T} = 0$;
% 		\For{each training pattern p}
% 		\State Calculate the output of the hidden and output layers (feedforward);
% 		\State Compute the error signals of the hidden and output layers;
% 		\State Adjust the weights of the output and hidden layers (backpropagate);
% 		\State $\epsilon_{T}$ += $[\epsilon_{T_{p}} = \sum^{K}_{k=1}(t_{k,p} - o_{k,p})^{2}]$;
% 		\EndFor
% 		\State $T += 1$;
% 		\EndWhile
% 	\end{algorithmic}
% \end{algorithm}

@@

@@

\subsubsection{Momentum}\label{sec:heuristics:gd:momentum}

@@

\subsubsection{Nesterov Accelerated Gradients}\label{sec:heuristics:nag}

@@

\subsubsection{Adaptive Gradients}\label{sec:heuristics:adagrad}

@@

\subsubsection{Adaptive Learning Rate}\label{sec:heuristics:adadelta}

@@

\subsubsection{Root Mean Squared Error Propagation}\label{sec:heuristics:rmsprop}

@@

\subsubsection{Adaptive Moments Estimation}\label{sec:heuristics:adam}

@@

\subsection{Meta-Heuristics}\label{sec:heuristics:mh}

@@

\subsubsection{Particle Swarm Optimisation}\label{sec:heuristics:mh:pso}

@@

\subsubsection{Differential Evolution}\label{sec:heuristics:mh:de}

@@

\subsubsection{Genetic Algorithms}\label{sec:heuristics:mh:ga}

@@
